---
title: "STA232B Lab 1"
author: "Po"
date: "1/6/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Homework 1

HW 1.1, HW 1.2, HW 1.3, Homework 1.4, HW 1.5, HW 1.6

#### 1.1
Write the equation (right above the question) into the from $\mathbf{y} = \mathbf{X}\boldsymbol{\beta} + \mathbf{Z}\boldsymbol{\alpha} + \boldsymbol{\epsilon}$. Write down $\mathbf{y}$, $\boldsymbol{\beta}$, $\boldsymbol{\alpha}$, $\boldsymbol{\epsilon}$, $X$ and $Z$ explicitly. 

You can use Kronecker product for $\mathbf{X}$ and $\mathbf{Z}$. For example, $\boldsymbol{\beta} = (\beta_1, \dots, \beta_5)$ and $\mathbf{X} = \mathbf{1}_m \otimes \mathbf{I}_5$.

#### 1.2 and 1.3

Write the equation into the form of ANOVA model. What are $Z_1, \dots, Z_s$?

#### 1.4

i) Use Total variance formula for part (i)  https://en.wikipedia.org/wiki/Law_of_total_variance
ii) Write the growth curve model in example 7 into the form of Longitudinal model. What are $\mathbf{y}, \mathbf{X}, \boldsymbol{\beta}, \mathbf{Z}_i, \boldsymbol{\alpha}_i, \mathbf{G}_i, \boldsymbol{\epsilon}_i, \mathbf{R}_i$? What variables are inside $\theta$?

#### 1.5

a) Write down $\mathbf{V}$ in terms of $\tau, \sigma^2_r, \mathbf{Z}_r$.
b) Write down $\mathbf{V}$ in terms of $\tau, \phi, \sigma_0, \sigma_1, \rho, \beta_{0i}, \beta_{1i}$.

#### 1.6

- Use formula for total covariance to show the equation. https://en.wikipedia.org/wiki/Law_of_total_variance
- To show normality, there are serveral ways. One way is to explicitly show that the likelihood function of the model is exactly an Gaussian one. Another way is to show the moment generating function is exactly the Gaussian one.
https://en.wikipedia.org/wiki/Multivariate_normal_distribution

# Computing

## `lme4`

```{r, eval=F}
library(lme4)
?lme4
?merMod
```

```{r}
methods(class="merMod")
```

```{r}
?getME
```

## Lamb data

Let's look at the Lamb dataset in the textbook (the table 1.2 in Jiang 2007, same as the table 1 in Harville 1985).

>    Harville, David A., and Alan P. Fenech. "Confidence Intervals for a Variance Ratio, or for Heritability, in an Unbalanced Mixed Linear Model." Biometrics 41, no. 1 (1985): 137-52. doi:10.2307/2530650. <br>
> https://www.jstor.org/stable/2530650

> For purposes of illustration, we introduce in Table 1 some data consisting of the weights at
> birth of 62 single-birth male lambs. These data (provided by G. E. Bradford, Department
> of Animal Science, University of California, Davis) come from five distinct population
> lines (two control lines and three selection lines). Each lamb was the progeny of one of 23
> rams, and each lamb had a different dam. Age of dam was recorded as belonging to one of
> three categories, numbered 1 (1-2 years), 2 (2-3 years), and 3 (over 3 years).

Sheep glossary: http://www.sheep101.info/201/glossary.html

`lamb_lab1.csv` contains a subset of the data (Line 1 and 2).

```{r}
library(lme4)
library(tidyverse)

lamb <- read.csv(file = "lamb_lab1.csv") %>%
  mutate_at(c("Sire", "Line", "Age"), as.factor)

lamb %>% 
  knitr::kable()
```

Let's fit a linear mixed effect model!
```{r}
lamb_fitted <- lmer(Weight ~ Line + Age - 1 + (1|Sire), data=lamb)
summary(lamb_fitted)
```

```{r}
VarCorr(lamb_fitted)
```

```{r}
getME(lamb_fitted, "X")
```

```{r}
getME(lamb_fitted, "Z")
```

```{r}
getME(lamb_fitted, "fixef")
```

```{r}
ranef(lamb_fitted)
```
