---
title: "STA232Lab5"
author: "Po"
date: "2/3/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Project 1

1. Reading: Write a few sentences to summary what is going on.
2. Data Entry: Input the data from the table in textbook so you can fit the model here in R. Find you way to do it.
3. Analysis using Maximum Likelihood: Fit the model with MLE, i.e. `REML = FALSE`.
   i. Asymptotic covariance matrix, a.k.a. textbook formula
   ii. Parametric bootstrap
4. Analysis using REML: Fit the model with MLE, i.e. `REML = TRUE`.
   i. Asymptotic covariance matrix, a.k.a. textbook formula
   ii. Parametric bootstrap
5. Discussion: Write something educational.

Don't hand in raw R output as your report. Write paragraphs and tables.

### Asymptotic covariance matrix

Study section 1.3 of the textbook carefully.

Long story short: Get the inverse of the Fisher information matrix of the $\hat{\theta}$'s. The recipe is section 1.3 of the textbook.

## `lme4`

- https://arxiv.org/pdf/1406.5823.pdf

- https://github.com/lme4/lme4

## Lamb data

Let's look at the Lamb dataset in the textbook (the table 1.2 in Jiang 2007, same as the table 1 in Harville 1985).

>    Harville, David A., and Alan P. Fenech. "Confidence Intervals for a Variance Ratio, or for Heritability, in an Unbalanced Mixed Linear Model." Biometrics 41, no. 1 (1985): 137-52. doi:10.2307/2530650. <br>
> https://www.jstor.org/stable/2530650

> For purposes of illustration, we introduce in Table 1 some data consisting of the weights at
> birth of 62 single-birth male lambs. These data (provided by G. E. Bradford, Department
> of Animal Science, University of California, Davis) come from five distinct population
> lines (two control lines and three selection lines). Each lamb was the progeny of one of 23
> rams, and each lamb had a different dam. Age of dam was recorded as belonging to one of
> three categories, numbered 1 (1-2 years), 2 (2-3 years), and 3 (over 3 years).

Sheep glossary: http://www.sheep101.info/201/glossary.html

`lamb_lab1.csv` contains a subset of the data (Line 1 and 2).

```{r message=FALSE}
library(lme4)

lamb <- read.csv(file = "lamb_lab1.csv")
cols <- c("Sire", "Line", "Age")
lamb[cols] <- lapply(lamb[cols], as.factor)

knitr::kable(head(lamb))
```

Let's fit a linear mixed effect model!
```{r}
lamb_fitted <- lmer(Weight ~ Line + Age - 1 + (1|Sire), data=lamb)
summary(lamb_fitted)
```

Confident interval
```{r}
confint(lamb_fitted, method = "Wald")
```

```{r}
confint(lamb_fitted, method = "boot")
```

```{r}
?confint.merMod
```

Fisher information matrix
```{r}
y = getME(lamb_fitted, "y")
X = getME(lamb_fitted, "X")
Z = getME(lamb_fitted, "Z")
n = getME(lamb_fitted, "n")

sigma_sq = sigma(lamb_fitted)
tau_sq   = unlist(VarCorr(lamb_fitted))

ZZt = Z %*% t(Z)
V = tau_sq * diag(n) + sigma_sq * ZZt
Vinverse = solve(V)
P = Vinverse - solve(V, X %*% solve(t(X) %*% solve(V, X), t(X)%*%Vinverse))

tr = function(x) sum(diag(x))

Fisher_informatin_matrix_beta = t(X)%*%solve(V, X)
Fisher_informatin_matrix_varc = 0.5*matrix(c(tr(P%*%P), tr(P%*%ZZt%*%P),
                                        tr(P%*%P%*%ZZt), tr(P%*%ZZt%*%P%*%ZZt))
                                      , nrow=2, ncol=2)
```

```{r}
solve(Fisher_informatin_matrix_beta)
```

```{r}
solve(Fisher_informatin_matrix_varc)
```

```{r, message=FALSE, eval=FALSE}
?getME
?VarCorr
```

Simulation with fitted parameters.
```{r}
simulate(lamb_fitted, nsim = 5, seed = 1997)
```

Lets do parametric bootstrap!
```{r}
mySumm <- function(mod) {
   c(sigma_e = sigma(mod), sigma_r = sqrt(unlist(VarCorr(mod))))
}
booted_lamb_fitted <- bootMer(lamb_fitted, mySumm, nsim = 100, seed = 2047)
```

```{r}
library(boot) # for nice print-out
booted_lamb_fitted
```

```{r}
?simulate.merMod
?bootMer
```

EBLUP of the "mean of lamb weights conditional on sire effects" and sire effects
```{r}
lamb_extend <- lamb
lamb_extend["EBLUP_Weight_way1"] <- predict(lamb_fitted)
lamb_extend["EBLUP_Weight_way2"] <- X%*%fixef(lamb_fitted) + as.matrix(Z%*%ranef(lamb_fitted)$Sire$`(Intercept)`)
lamb_extend["EBLUP_Zalpha"] <- as.matrix(Z%*%ranef(lamb_fitted)$Sire$`(Intercept)`)
knitr::kable(head(lamb_extend))
```

```{r}
?predict.merMod
```
